Functory: Distributed Computing for the Common Man

### 2010-07-02 doc/icfp-2010/main.tex
submission to ML workshop

\titlebanner{submitted to ICFP 2010}        % These are ignored unless
\preprintfooter{Experience Report: Distributed Computing for the Common Man}   % 'preprint' option specified.

\title{Experience Report: Distributed Computing for the Common Man}
%\subtitle{Subtitle Text, if any}

>>>>

\titlebanner{Technical Report}        % These are ignored unless
\preprintfooter{Functory: Distributed Computing for the Common Man}   % 'preprint' option specified.

\title{Functory: Distributed Computing for the Common Man}
%\subtitle{Subtitle Text, if any}

### 2010-07-02 doc/ml-workshop-2010/main.tex
submission to ML workshop

\titlebanner{submitted to Workshop on ML 2010}        % These are ignored unless
\preprintfooter{Distributed Computing for the Common Man}   % 'preprint' option specified.

\title{Distributed Computing for the Common Man}
%\subtitle{Subtitle Text, if any}

@@

This paper introduces a generic library for distributed computing. This
work was initially motivated by the computing needs that exist in our
own research team. Our applications include large-scale deductive
program verification, which amounts to checking the validity of a
large number of logical formulas using a variety of automated theorem
provers~\cite{filliatre07cav}. Our computing infrastructure consists
of a few powerful multi-core machines (typically 8 to 16 cores) and
several desktop PCs (typically dual-core). However, for our
application needs, there is no library that helps in exploiting such a
computing infrastructure in our favorite functional programming
language.  Hence we designed and implemented such a
library\footnote{This library is implemented in \Ocaml, but the
  implementation should be straightforward in any functional
  programming language.}, which is the subject of this paper.

The distributed computing library presented in this paper is not a
library that helps in parallelizing computations. Rather, it provides
facilities for reliable, distributed execution of parallelizable
computations. In particular, it provides a set of APIs that allows the
execution of large-scale parallelizable computations, very relevant to
our application needs (and presumably relevant to a variety of
real-world applications), over multiple cores in the same machine or
over a network of machines. 
The most important features of our library are the following:
\begin{itemize}
\item \emph{Genericity}: 
  it allows various patterns of polymorphic computations;
\item \emph{Simplicity}: switching between multiple cores on the same
  machine and a network of machines is as simple as changing a couple
  of lines of code;
\item \emph{Task distribution and fault-tolerance}: 
  it relieves the user from implementing task distribution routines
  and reliable fault-tolerance mechanisms.
\end{itemize}
It is worth noting that the library is not targeted at applications
running on server farms, crunching enormous amounts of data. Neither
is it limited to research labs possessing few powerful
machines. It serves a variety of users and a wide
spectrum of needs, from desktop PCs to networks of machines, and hence
the title.

@@

\paragraph{N-queens.} 
The following table shows execution times\footnote{machine
  characteristics: 8 cores Intel Xeon 3.2 GHz running Debian Linux.} 
in seconds for various values of $N$
and our three different implementations (sequential, 8 cores on a
single machine and a network of 8 workers).
\begin{center}
  \begin{tabular}{|r|r|r|r|r|}
    \hline
    N & \#tasks  & sequential& cores                 & network 
    \\\hline\hline
    16 &   16    &  15.2     &   2.04 (7.45$\times$) &  2.35  (6.47$\times$) 
    \\\hline
       &  210    &  15.2     &   2.01 (7.56$\times$) & 21.80  (0.69$\times$)
    \\\hline
    17 &   17    & 107.0     &  17.20 (6.22$\times$) & 16.20  (6.60$\times$)
    \\\hline
       &  240    & 107.0     &  14.00 (7.64$\times$) & 24.90  (4.30$\times$)
    \\\hline
    18 &   18    & 787.0     & 123.00 (6.40$\times$) & 125.00 (6.30$\times$)  
    \\\hline
       &  272    & 787.0     & 103.00 (7.64$\times$) & 124.00 (6.34$\times$)  
    \\\hline
    19 &   19    &6120.0     & 937.00 (6.53$\times$) & 940.00 (6.51$\times$)  
    \\\hline
       &  306    &6130.0     & 796.00 (7.70$\times$) & 819.00 (7.48$\times$)
    \\\hline
  \end{tabular}
\end{center}
% From the table above, it is clear that the \of{Cores} and \of{Network}
% implementations provide a significant speedup. As evident from the
% last row, the speedup is almost 8, which is also the number of
% cores we use.  It is also evident from the last column that the
% \of{Network} implementation performs significantly better when the
% computation time dominates in the total execution time.  The two extreme
% cases correspond to the second and the last row: in the second row, the
% communication time dominates and is in fact more than 91\%\ of the
% total execution time; on the other hand, for the last row
% communication time amounts to just 4.6\%\ of the total execution time.
% As expected, the network implementation is only beneficial when the
% computation time for each individual task is significant, which is the
% case in realistic examples.

@@

>>>>

\titlebanner{submitted to Workshop on ML 2010}        % These are ignored unless
\preprintfooter{Functory: Distributed Computing for the Common Man}   % 'preprint' option specified.

\title{Functory: Distributed Computing for the Common Man}
%\subtitle{Subtitle Text, if any}

@@

This paper introduces a generic library for distributed computing. This
work was initially motivated by the computing needs that exist in our
own research team. Our applications include large-scale deductive
program verification, which amounts to checking the validity of a
large number of logical formulas using a variety of automated theorem
provers. Our computing infrastructure consists
of a few powerful multi-core machines (typically 8 to 16 cores) and
several desktop PCs (typically dual-core). However, for our
application needs, there is no library that helps in exploiting such a
computing infrastructure in our favorite functional programming
language.  Hence we designed and implemented such a
library\footnote{This library is implemented in \Ocaml, but the
  implementation should be straightforward in any functional
  programming language.}, which is the subject of this paper.

The distributed computing library presented in this paper is not a
library that helps in parallelizing computations. Rather, it provides
facilities for reliable, distributed execution of parallelizable
computations. In particular, it provides a set of APIs that allows the
execution of large-scale parallelizable computations, very relevant to
our application needs (and presumably relevant to a variety of
real-world applications), over multiple cores in the same machine or
over a network of machines. 
The most important features of our library are the following:
\begin{itemize}
\item \emph{Genericity}: 
  it allows various patterns of polymorphic computations;
\item \emph{Simplicity}: switching between multiple cores on the same
  machine and a network of machines is as simple as changing a couple
  of lines of code;
\item \emph{Task distribution and fault-tolerance}: 
  it relieves the user from implementing task distribution routines
  and fault-tolerance mechanisms.
\end{itemize}
It is worth noting that the library is not targeted at applications
running on server farms, crunching enormous amounts of data. Neither
is it limited to research labs possessing few powerful
machines. It serves a variety of users and a wide
spectrum of needs, from desktop PCs to networks of machines, and hence
the title.

@@

\paragraph{N-queens.} 
The following table shows execution times\footnote{machine
  characteristics: 8 cores Intel Xeon 3.2 GHz running Debian Linux.} 
in seconds for various values of $N$
and our three different implementations (sequential, 8 cores on a
single machine and a network of 8 workers).
\begin{center}
  \begin{tabular}{|r|r|r|r|r|}
    \hline
    N & \#tasks  & sequential& cores                 & network 
    \\\hline\hline
    16 &   16    &  15.2     &   2.04 (7.45$\times$) &  2.35  (6.47$\times$) 
    \\\hline
       &  210    &  15.2     &   2.01 (7.56$\times$) & 21.80  (0.69$\times$)
    \\\hline
    17 &   17    & 107.0     &  17.20 (6.22$\times$) & 16.20  (6.60$\times$)
    \\\hline
       &  240    & 107.0     &  14.00 (7.64$\times$) & 24.90  (4.30$\times$)
    \\\hline
    18 &   18    & 787.0     & 123.00 (6.40$\times$) & 125.00 (6.30$\times$)  
    \\\hline
       &  272    & 787.0     & 103.00 (7.64$\times$) & 124.00 (6.34$\times$)  
    \\\hline
    19 &   19    &6120.0     & 937.00 (6.53$\times$) & 940.00 (6.51$\times$)  
    \\\hline
       &  306    &6130.0     & 796.00 (7.70$\times$) & 819.00 (7.48$\times$)
    \\\hline
  \end{tabular}
\end{center}
As evident from the last row, the speedup almost reaches 8, which is
also the number of cores we use.

% From the table above, it is clear that the \of{Cores} and \of{Network}
% implementations provide a significant speedup. As evident from the
% last row, the speedup is almost 8, which is also the number of
% cores we use.  It is also evident from the last column that the
% \of{Network} implementation performs significantly better when the
% computation time dominates in the total execution time.  The two extreme
% cases correspond to the second and the last row: in the second row, the
% communication time dominates and is in fact more than 91\%\ of the
% total execution time; on the other hand, for the last row
% communication time amounts to just 4.6\%\ of the total execution time.
% As expected, the network implementation is only beneficial when the
% computation time for each individual task is significant, which is the
% case in realistic examples.

@@

\section{Conclusion}

We presented \functory, a distributed programming environment for \Ocaml.
The main features are the genericity of the interface,
which makes use of polymorphic higher-order functions, and the ability
to easily switch between sequential, multi-core, and network
implementations. A distinguishing feature of our library is a
robust fault-tolerance mechanism which relieves
the user of cumbersome implementation details.

The most relevant application domains for this library are the ones
where computation time is much greater than communication time, which
is the case in many real-world applications. This results in a speedup
ratio which is close to the total number of cores.
